---
title: "Houston analysis"
format: html
editor: visual
author: "Austin Martinez"
date: "12/01/2025"
execute:
  message: false
  warning: false
---

# **Purpose**

The purpose of this project was to analyze the extent of the power outage that took place in Houston, Texas due to the winter storms in February 2021. Socioeconomic data from the U.S. Census Bureau was used estimate the number of houses that had a power outage and evaluate whether communities that had socioeconomically differences were effected differently.

```{r}
# Loading in Libraries
library(tmap)
library(here)
library(dplyr)
library(sf)
library(ggplot2)
library(terra)
library(stars)
library(knitr)
library(kableExtra)

```

# **Reading in Data**

```{r}
# Read in the four raster tiles for Night lights data
vnp_05 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
vnp_06 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
vnp_05_2 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
vnp_06_2 <- read_stars(here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

```{r}
# Read in data for excluding highways from the cropped blackout mask
roads <- st_read(here("data", "gis_osm_roads_free_1.gpkg"), query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'", quiet = TRUE)
```

```{r}
# Read in data for identifying the number of homes likely impacted by blackouts
buildings <- st_read(here("data", "gis_osm_buildings_a_free_1.gpkg"),
                     query = "SELECT *
                              FROM gis_osm_buildings_a_free_1
                              WHERE (type IS NULL AND name IS NULL)
                                 OR type IN ('residential','apartments','house','static_caravan','detached')", quiet = TRUE)

```

```{r}
#| message: false
# read in Socioeconomic data
acs_geom <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                    layer = "ACS_2019_5YR_TRACT_48_TEXAS", quiet = TRUE)|>
                    dplyr::select(-c(GEOID)) |> 
                    rename(GEOID = GEOID_Data) |> 
                    st_transform(crs = 3083)

acs_income <- st_read("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb",
                      layer = "X19_INCOME", quiet = TRUE) |> dplyr::select(c(GEOID, B19013e1, B19013m1))

```

# **Data Cleaning**

### Create blackout mask

#### To identify places that experienced a blackout, a “mask” was created that indicates for each cell whether or not it experienced a blackout.

To do so, the change in night lights intensity (presumably) caused by the storm was found by creating a raster object for each day (2021-02-07 and 2021-02-16). The difference raster was then reclassified such that any location that experienced a drop of more than 200 nW cm-2sr-1 experienced a blackout. NAs were assigned to all locations that experienced a drop of less than 200 nW cm-2sr-1 change. The mask then vectored using st_as_sf() and any invalid geometries were fixed with st_make_valid(). The mask was then cropped to the Houston area using following coordinates: (-96.5, 29), (-96.5, 30.5), (-94.5, 30.5), (-94.5, 29). The cropped mask data set then needed to be re-projected to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area).

```{r}
# Merge 2021-02-07 data (before the storm)
before <- st_mosaic(vnp_05, vnp_06)

# # Merge 2021-02-16 data (after the storm)
after <- st_mosaic(vnp_05_2, vnp_06_2)

# Find the change in night light
change_in_light <- before - after

# Identify blackout areas
blackout_mask <- change_in_light > 200

# Assign NAs to all locations that experienced a drop of less than 200
blackout_mask[blackout_mask <= FALSE] <- NA
blackout_mask[blackout_mask > TRUE]  <- 1

# Convert raster mask to vector polygons
blackout_vec <- st_as_sf(blackout_mask, as_points = FALSE, merge = TRUE, na.rm = TRUE)

# Make polygons valid
blackout_vec <- st_make_valid(blackout_vec)
```

```{r}
#| message: true
# Crop to the Houston area
houston_box <- st_bbox(c(xmin = -96.5, ymin = 29, xmax = -94.5, ymax = 30.5), crs = st_crs(blackout_vec))

# Convert to simple feature polygon
houston_box <- st_as_sfc(houston_box)

# Crop to the Houston bounding box
houston_blackout <- st_crop(blackout_vec, houston_box)

# Transform houston_blackout CRS to EPSG:3083
houston_blackout_3083 <- st_transform(houston_blackout, crs = 3083)

# Check if CRSs is EPSG:3083
if (st_crs(houston_blackout_3083)$epsg == 3083) {
  message("🟢 CRS is EPSG:3083")
} else {
  warning("❗ CRS is NOT EPSG:3083")
}
```

### Exclude highways from the cropped blackout mask

#### Any locations within 200 meters of all highways in the Houston area were excluded to ensure accuracy. This is because highways may have experienced changes in their night light intensities that were unrelated to the storm.

To accomplish this the CRS of `roads` was transformed to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area) and areas within 200m of all highways were identified using a buffer. The highway buffers were merged into a single polygon. This polygon was then used to find areas that experienced blackouts that are further than 200m from a highway.

```{r}
# Transform roads CRS to EPSG:3083
roads_3083 <- st_transform(roads, crs =  3083)

# Buffer highways by 200 meters
highway_buffer <- st_buffer(roads_3083, dist = 200)

# Merge all highway buffers into a single polygon
highway_union <- st_union(highway_buffer)

# Remove highway areas from blackout polygons
blackout_no_highway <- st_difference(houston_blackout_3083, highway_union)

```

### Identify the number of homes likely impacted by blackouts

To identify homes that overlap with areas that experienced blackouts the CRS of `buildings` was changed to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area) and a spacial intersection was done on `buildings_3083` and `blackout_no_highway`.

```{r}
# Transform buildings CRS to EPSG:3083
buildings_3083 <- st_transform(buildings, crs =  3083)

# Preform intersection on `buildings_3083` and `blackout_no_highway`
affected_buildings <- st_join(buildings_3083, blackout_no_highway, join = st_intersects, left = FALSE)
```

## Socioeconomic

Since getting socioeconomic information for every home isn't doable, data from the U.S. Census Bureau’s American Community Survey for census tracts in 2019 was instead obtained. The folder ACS_2019_5YR_TRACT_48.gdb is an ArcGIS “file geodatabase”, a multi-file proprietary format that’s roughly analogous to a GeoPackage file.

To get a feature layer that sf can use, the geometry layer (ACS_2019_5YR_TRACT_48_TEXAS) and the income layer (X19_INCOME) were combined using a left join on the GEOID column. The CRS was then transformed to EPSG:3083 (NAD83 / Texas Centric Albers Equal Area) to match the CRS of the blackout and building data.

```{r}
# Check layer names
st_layers("data/ACS_2019_5YR_TRACT_48_TEXAS.gdb")
```

```{r}
# Preform a left join on the GEOID column
acs_full <- left_join(acs_geom, acs_income, by = "GEOID")

# Transform acs_full CRS to EPSG:3083
acs_full_3083 <- st_transform(acs_full, 3083)
```

# **Data Cleaning**

### A set of maps comparing night light intensities before and after the first two storms

```{r}
# Apply log(1 + x) to make before visible
before_log <- log1p(before)

# Plot a map comparing night light intensities before two storms
tm1 <- tm_shape(before_log) +
  tm_raster(
    style = "cont",
    palette = "viridis",
    title = "Light Intensity nW cm^-2 sr^-1"  ) +   
  tm_layout(
    main.title = "Light Intensity Before the Storm (2021-02-07)",
    legend.outside = TRUE
  )  +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass()

# Show the map
tmap_arrange(tm1)
```

```{r}
# Apply log(1 + x) to make after visible
after_log  <- log1p(after)

# Plot a map comparing night light intensities after two storms
tm2 <- tm_shape(after_log) +
  tm_raster(
    style = "cont",
    palette = "viridis",
    title = "Light Intensity nW cm^-2 sr^-1") + 
  tm_layout(
    main.title = "Light Intensity After the Storm (2021-02-07)",
    legend.outside = TRUE ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass()

# Show the map
tmap_arrange(tm2)


```

### A map of the homes in Houston that lost power

```{r}
# Plot a map of the homes in Houston that lost power
tm_homes <- tm_shape(affected_buildings) +
  tm_basemap("OpenStreetMap") +
  tm_bubbles(
    size = 0.2,
    col = "red"
    ) +
  tm_add_legend(
    type = "symbol",
    labels = "Affected Homes",
    col = "red",
    size = 0.5
  ) +
  tm_layout(
    main.title = "Homes Affected by Blackouts in Houston, TX",
    legend.outside = TRUE
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass()

# Show the map
tm_homes

```



```{r}
# Count rows to find the number of homes that experienced a blackout
num_affected_homes <- nrow(affected_buildings)

# Create a data frame with the count of homes that experienced a blackout
affected_summary <- data.frame( City = "Houston", Homes_Affected = num_affected_homes)

# Display number of homes that experienced a blackout with a kable
kable(affected_summary,
caption = "Estimate of the Number of Homes in Houston that Lost Power") |>
kable_styling(full_width = FALSE, bootstrap_options = "bordered")


```





### A map of the census tracts in Houston that lost power


```{r}
# perform an intersection on acs_full_3083 and blackout_no_highway
houston_acs_blackout <- st_join(acs_full_3083, blackout_no_highway, join = st_intersects, left = FALSE)

# Create a boolean column to show which had/didn't have a power outage
acs_full_3083 <- acs_full_3083 %>%
  mutate(blackout = GEOID %in% houston_acs_blackout$GEOID)

# Make a map of a map of the census tracts in Houston that lost power
tm_blackout_tracts <- tm_shape(acs_full_3083) +
  tm_polygons(
    col = "blackout",
    palette = c("grey", "red"),
    title = "Blackout Status"
  ) +
  tm_layout(
    main.title = "Census Tracts Affected by Winter Storm Uri in Houston",
    legend.outside = TRUE
  ) +
  tm_scale_bar(position = c("left", "bottom")) +
  tm_compass()

# Show the map
tm_blackout_tracts

```

```{r}
# Make blackout_mask into polygons
blackout_vec <- st_as_sf(blackout_mask, as_points = FALSE, merge = FALSE, na.rm = TRUE)

# Change blackout_vec's CRS to 3083
blackout_vec_3083 <- st_transform(blackout_vec, 3083)

# Check which census tracts intersect a blackout polygon
tracts_with_blackout <- st_intersects(acs_full_3083, blackout_vec_3083)
acs_full_3083$blackout <- lengths(tracts_with_blackout) > 0

# Make a box plot comparing the distributions of median household income for census tracts that did and did not experience blackouts
ggplot(acs_full_3083, aes(x = blackout, y = B19013e1, fill = blackout)) +
  geom_boxplot() +
    scale_fill_manual(
    values = c("TRUE" = "gray", "FALSE" = "yellow"),
    labels = c("TRUE" = "Blackout", "FALSE" = "No Blackout"), # Change value names in legend
    name = "Blackout Status"
  ) +
  scale_x_discrete(labels = c("No Blackout", "Blackout")) + # Change value names below box plots
  labs(
    title = "Median Household Income by Blackout Status",
    x = "Blackout Status",
    y = "Median Household Income (USD$)"
  ) +
  theme_minimal()


```



### A brief reflection (approx. 100 words) summarizing your results and discussing any limitations to this study

After this analysis was conducted it was found that the estimated number of homes in Houston, Texas that experienced a power outage due to the Winter storm Uri in Febuary 2021 was 157,417. A big limitation of this study was the data. Since the blackout mask was only based on two days of night light data, its probable that not all outages were captured. Another limitation was how the study was conducted. St_intersection was used, which assumes houses partially in the blackout zone experienced a power outage. This might have caused an over estimation of the extent of the blackout.


#### **References**

-   Román, M. O., Wang, Z., Sun, Q., Kalb, V. L., Miller, S. D., Molthan, A., Schultz, L., Bell, J., Stokes, E. C., Pandey, B., et al. (2018). NASA’s Black Marble nighttime lights product suite (VNP46). Remote Sensing of Environment, 210, 113–143. <https://doi.org/10.1016/j.rse.2018.03.017> Retrieved November 10

-   OpenStreetMap Contributors. (2025). OpenStreetMap database [Data set]. Geofabrik GmbH. Retrieved November 10, 2025

-   U.S. Census Bureau. (2020). TIGER/Line shapefiles and American Community Survey 2019 (5-year estimates), Texas — Census tract level (ACS_2019_5YR_TRACT_48_TEXAS) [Data set]. U.S. Department of Commerce. Retrieved November 10, 2025

-   Lee, C. C., Maron, M., & Mostafavi, A. (2022). Community-scale big data reveals disparate impacts of the Texas winter storm of 2021 and its managed power outage. Humanities and Social Sciences Communications, 9(1), 335. <https://doi.org/10.1057/s41599-022-01353-8> Retrieved November 10, 2025